\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[backgroundcolor=yellow]{todonotes}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{tikz}
\usepackage{rotating}
\usepackage{url}
\usepackage{listings}

\usetikzlibrary{matrix,shapes,arrows,positioning,chains, calc}
 
\title{Analysys of SQRL Authentication Protocol using proverif/spass/whatever}
\author{Dettoni, Claudio\\
	\texttt{cd611@cam.ac.uk}
	\and
	Perin, Lucas\\
	\texttt{lucas.perin@posgrad.ufsc.br}
}
\date{December 2016}

\begin{document}

\maketitle

\begin{abstract}
	In this paper we give a simplified description of the SQRL authentication
	protocol and how the relay attack works against it. Numerous other
	protocols that classify in the same family of visual codes authentication protocols
	are also vulnerable to this attack. We run an experiment for the formal verification of 
	this family of protocols using Spass, attempting to detect the attack with an
	automated tool.
\end{abstract}

\section{Intoduction}
	Authentication protocols play an important role in modern internet
	usability. Most social network providers already have implemented
	some sort of authentication method to make it easy for users to connect
	to different services using the same account and credentials. The Single Sign On 
	(SSO) protocols implemented by Google using OpenID \cite{openid} is a
	fine examples of this. However, we know that the good and 
	old password is no longer reliable and has poor usability \cite{password}.
	Furthermore, most services on the web still use pass phrases for authentication,
	since this has been the main method used in previous decades. Hence,
	a common aproach to provide more security to personal accounts has 
	been the adoption of two-step verification, using and One Time Passwords (OTP),
	provided by a trusted mobile app or token. This	method has gained some 
	space in authentication providers such as the
	one previously mentioned. Therefore, we now have a mixed authentication process
	where the user still uses his passwords, but can be prompted for an
	OTP optionally.

	Numerous authentication protocols have been proposed in the past
	decade with the goal of replacing pass phrases with something that
	is actually secure and reliable for authentication providers. One
	recent example is the Pico project \cite{pico}, with the main goal of
	completely removing passwords. An other example of similar protocols
	are tiQR \cite{tiqr} and snap2pass \cite{snap2pass}.
	However, we have yet to see the adoption of one of these methods
	by a popular authentication provider to lead its use across the
	Internet. 

	In this paper, we analyse the SQRL protocol \cite{sqrl}. This protocol
	has drawed attention since its publication due to its simplicity. However,
	simple does not mean secure. In \cite{relayattack} the authors
	describe how to use relay attacks to break through the SQRL authentication
	protocol. This attack can also be used with other authentication
	protocols that use images for authentication, such as the previously mentioned
	Pico, tiQR and snap2pass.
	Our goal is to use Spass, a tool for formal verification of cryptgraphic
	protocols, to detect the relay attack in these protocols in an automated fashion.

\section{Secure Quick Reliable Login Protocol}

	The SQRL protocol, short for Secure Quick Reliable login, is an alternative method 
	for authentication replacing user names and passwords. It uses a QR code
	similar to other protocols, such as the Pico Project and TiQR. To log in
	the user should have a device that is capable of storing offline cryptographic
	keys and compute hash and cryptographic operations. In this work, we will consider that
	this is an application runing in an mobile device, namely the SQRL App. 
	Additionally, denote an existing authentication service provider as ASP. Now
	we can describe the main phases of the protocol as the \emph{Setup, Identity 
	Creation} and \emph{Authentication} phases.

	The \emph{Setup} phase is a standart cryptographic setup procedure where the
	SQRL App generates a random 256 bit master secret (MS). This secret is stored 
	offline and 
	is never to be shared. The following phases will always be derivated of this
	secret, but never give it away. The \emph{Identity Creation} phase and the
	\emph{Authentication} phase are operations that should be syncronized with
	each ASP the user wants to authenticate to. In other words, the user will
	have to create an identity for each ASP and each ASP is responsible 
	for authenticating the user's credential upon log in.

	The \emph{Identity Creation} phase is performed when the user wishes to create an
	identity for a specific ASP. In this step he will create a Domain Secret (DS)
	derivated from the service's domain address and the MS, by using an HMAC function.
	This DS is used only for this specific ASP and no others. Furthermore, the
	MS should remain stored safely in case the user needs to create a new DS. In 
	this same phase, a public key (KU) and a private key (PK) are derived from the
	DS, where the KU is then forwarded to the ASP.


	The \emph{Authentication} phase should be the operation that users use the
	most. After the Setup and Identity Creation, the user should have a KU and
	a PK for each of his ASPs. To authenticate the user, the ASP generates a 
	nonce and displays it as a QR code for the user to scan. The user scans the
	nonce using the SQRL App and signs it with the PK. The signature is then
	sent to the ASP from the SQRL APP using a different communication channel,
	that is, a secure connection directly from the SQRL App and the ASP.
	Finally, the ASP verifies the signature received from the SQRL app and 
	approves or rejects the user's login attempt. See Figure \ref{fig:sqrlauth}

	\input{diag1}

\subsection{Cryptographic Algorithm Remarks}
	The underlaying cryptographic algorithms used to generate random numbers, derivate
	key pairs from master secret and signature schemes are not detailed in this work.
	These are detailed and publicly availabe in the protocol's website (REF) and should not
	affect the attack we analyse here. 

\subsection{Security Remarks}
	The SQRL protocol, when used for website authentication, relies heavily on
	the \emph{Transport Layer Security} (TLS) cipher suite. The TLS ensures that
	the protocol phases should be secured with integrity, privacy and authenticity
	to prevent attacks such as the \emph{Man-in-the-middle} attack. A more extensive 
	analysis on the security of SQRL, can be foundin (REF) as well as in discussion 
	groups in the \emph{stack exchange} website \cite{sqrlreport}.

%\section{Similar Protocols}
%\todo[inline]{OpenID, Pico, TiQR}

\section[Relay attack to visual code authentication schemes]{Relay attack to visual code
authentication \\ schemes}
	The relay attack, as describe by the authors in \cite{relayattack}, is not a specific
	attack to the SQRL protocol, but rather to a family of protocol based on
	visual codes. Let us consider then a generic protocol that is compatible
	with the SQRL Authentication phase and other protocols in the same family.
	See Figure \ref{fig:genauth}.

	\input{diag2}
	
	In Steps 1 and 2, the ASP displays a nonce and the domain address in
	a visual code (QR code in the case of SQRL protocol). The viasual code is
	scanned by the user or transferred in some way to an authentication device.
	This device could be the user's browser, a desktop application or a mobile
	application, for example. The authentication device produces a proof of
	his credentials using the domain address and the nonce and sends it directly
	to the ASP. Lastly, the ASP validades the proof. Most importantly, in Step
	7, the ASP validates the user's current session. That is, if the user is
	logging into a website, the user's browser receives an authenticated session
	from the ASP as soon as the ASP accepts the proof sent by the authentication
	devide. The rellay attack on visual code authentication protocols based on Figure
	\ref{fig:genauth} is only possible for this specific reason. In the following,
	Figure \ref{fig:attackAuth} ilustrates the relay attack.

	\input{diag3}

	The relay attack is performed when and attacker successfully prompts and
	convinces the user to scan a visual code. By contrasting Figures \ref{fig:genauth} 
	and \ref{fig:attackAuth}, there are minimal changes. Supose the attacker
	knows an ASP where a given user has already created an identity (phase 2
	from SQRL protocol, for example). Then the attacker requests a code for himself 
	from this ASP and relays his visual code in such a way that the user is 
	convinced to scan it.
	This can be done in a similar way to phishing attacks, by presenting fake web
	pages or even taking over the user's computer. Once the visual code is
	scanned by the user, the authentication device will send a proof to the ASP.
	Lastly, the ASP recognizes and verifies the proof, however, in this case
	the authenticated session is sent to the attacker instead. 

	Different from
	phishing attacks, the relay attack does not steal user credentials like
	username or password. If the attack is successfull, the attacker will be
	authenticated by the user but will have no access to the user's secrets. 
	Hence, to attempt a new login, the attacker would have to repeat the attack
	for the same user.

	From the same work where the attack is published, the authors describe
	a possible solution to avoid it. The main idea of this proposal is to
	remove the step where the ASP validates the user's session.

	\input{diag4}
	
	The modified protocol now relies in a safe communication channel between
	the authentication device and the user's device. The nonce is now absent
	and the authenticated session is sent to the authentication device. Thus,
	if the attacker attempts a relay attack on this protocol, the authenticated
	session is no longer sent to him, but to the user's device.

\section{Spass implementation}
\subsection{Overview of protocol verification with Spass}
	As shown in \cite{analysis}, 
	protocols can be interpreted using first-order logic. In this case, every message in the protocol
	is translated to logic assertions as well as the attacker capabilities, and then try to prove that the
	attacker cannot achieve some state that would lead to a failure.

	The parties knownledge are defined as logic predicates. For example, if party \emph{a} knows the message
	\emph{m}, then the predicate $Knows(a,m)$ is true. This same predicate could also be described, less 
	generically, as $Aknows(m)$ for example, but this does not make a difference in logical expressiveness.

	To represent the communication channel, the messages are translated to predicates meaning that they were
	sent and implications that manipulations to the knowledge of the parties. To emulate the parties behaviour,
	for example answering a message, the universal operator is used. The following example shows how a very simple
	protocol would be represented using first order logic. 

	$$ (i) A \rightarrow B:m  $$
	$$ (ii) B \rightarrow A:m $$

	In this simple procotol, A send a message to B that responds the same message. Using first-order logic the protocol
	would be written as the following:

	$$ sent(a,b,m) $$
	$$ \forall xa, xm \bullet  sent(xa,b,xm) \implies Knows(b, xm) \land  sent(b, xa, xm)$$

	The first assertion kick-starts the protocol, with $A$ sending the first message. The second assertion mimics $B$ 
	behaviour when receiving some message, which is replying the same message.

	To represent the attacker capabilities, one can say that the attacker can read anything in the channel,
	
	$$ \forall xa, xb, xm \bullet  sent(xa,xb,xm) \implies Knows(attacker, xm) $$

	or the attacker can send messages to anyone
	
	$$ \forall x, xm \bullet  Knows(attacker, xm) \implies sent(attacker, x, xm) $$

	To prove something regarding the protocol, we need to express this as a logical predicate and try to derive it from
	the described rules. For example, in the previously described protocol, does the attacker know $m$? This is true
	because:
		
	\begin{align*}
	\shortintertext{By definition}
	&sent(a,b,m)\\
	&\forall xa, xb, xm \bullet  sent(xa,xb,xm) \implies Knows(attacker, xm)\\
	\shortintertext{Thus}
	&sent(a,b,m) \implies Knows(attacker, m)\\
	\shortintertext{By modus ponens}
	&Knows(attacker, m)\\
	\end{align*}

	Using this simple idea of logic predicates, more sophisticated protocols can be described. After fully writing them,
	one can try to prove more useful predicates, for example $Knows(attacker, key)$. In the case of complicated protocols
	there will be many logical assertions and it might be unfeasible to make proofs manually. That is why Spass is useful as it
	is an automated first-order logic prover.

\subsection{Visual code scheme implemented in Spass}
	Based on the protocol shown in figure 2, we will then write the protocol in the form explained above using first-order logic:
	
	\begin{align*}
	\shortintertext{If any party receives some data, they know this data}
	&\forall xa, xb, xm \bullet sent(xa,xb,xm) \implies knows(xb,xm) \\
	\shortintertext{The user starts the protocol, requesting the web page}
	&sent(user,asp,request(user))\\
	\shortintertext{For any request, the ASP replies with a code that is unique to that request}
	&\forall xa,xreq\bullet  sent(xa,asp,xreq) \implies sent(asp, xa, code(xreq))
	\shortintertext{The user, after receiving a code, sends to the scanner}
	&\forall xa, xcode \bullet sent(xa,user,xcode) \implies sent(user, scanner(user), xcode)\\
	\shortintertext{The scanner validates the code, and sends to the server}
	&\forall xu, xcode \bullet sent(xu,scanner(xu),xcode) \implies sent(scanner(xu), asp, sign(xcode, scanner(xu)))\\
	\shortintertext{The server then sends a cookie corresponding to the signature, to the user who requested}
	&\forall xusign, xureq \bullet sent(x,asp,sign(code(request(xureq)), scanner(xusign))) \implies sent(asp, xureq, cookie(xusign))\\
	\end{align*}

	Also, we describe the attacker capabilities:	
	
	\begin{align*}
	\shortintertext{The attacker can trick users to read arbitrary codes}
	&\forall xu, xm \bullet knows(attacker,xm) \implies sent(attacker,xu,xm) \\
	\shortintertext{The attacker can request pages to the ASP}
	&sent(attacker,asp,request(attacker))\\
	\end{align*}


	The mentioned logic was implemented and executed on Spass. We used it to prove 
	\footnote{The proofs produced by Spass are not very human-readable. We use an open 
	source tool to digest the result. It is available at 
	\url{https://github.com/dettonijr/spass-proof-parser}} 
	the following that the user can receive a cookie for 
	himself $knows(user,cookie(user))$. We also used it to prove a less desirable 
	feature: $knows(attacker,cookie(user))$.

	Similarly, we implemented the version mentioned in \ref{fig:proposal}. In this case
	the assertion $knows(user,cookie(user))$ can be proved, but $knows(attacker,cookie(user))$
	cannot.


\section{Final Remarks}
	It is worth noticing how the theorem prover can implicitly perform operations
	that are essentially very similar to human reasoning. For example, in the 
	axioms we defined that the attacker can send any message that they know to the user,
	also the user will scan any message that they receive. Using those two properties,
	the prover can automatically derive that the attacker can send any message they know
	to the scanner, even if this was not directly defined:
\lstset{basicstyle=\ttfamily\footnotesize,breaklines=true}
\begin{lstlisting}[frame=single]
Method: General Resolution
5	 Knows(attacker,U) -> M(sent(attacker,V,U))
7	 M(sent(U,user,V))*+ -> M(sent(user,scanner(user),V))
##
14	 Knows(attacker,U) -> M(sent(user,scanner(user),U))
\end{lstlisting}

	Our experiment's result demonstrate that the relay attack can be detected
	with the help of automated theorem provers. 
	Inspired by this, we can now build a tool using spass to
	help protocol developers to detect the relay attack without too much
	effort. In addition, the same analysis can be done to detect the relay
	attack in new authentication protocols or variations to come.

\section{Appendix}
\appendix

\section{Spass code}
\lstset{basicstyle=\ttfamily\footnotesize,breaklines=true}

\lstinputlisting[frame=single]{spass/simple.dfg}

\lstinputlisting[frame=single]{spass/corrected.dfg}

\section{Proofs}

\lstinputlisting[frame=single]{spass/proof-knows-user.txt}
\lstinputlisting[frame=single]{spass/proof-knows-attacker.txt}

\bibliographystyle{abbrv}
\bibliography{ref.bib}

\end{document}
